Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
non-termination for nested loops

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
insufficient for prove termination for many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

@@

\keywords
Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
Floating-Point Ranking Functions.

@@

When surveying the area of program termination chronologically, we observe
an initial focus on monolithic approaches based on a single measure shown to
decrease over all program
paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
recent techniques that use termination arguments based on Ramsey's
theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
Briefly, the transition relation is shown to be disjunctively well founded
by finding a set of simple functions, each of which proves well foundedness
for part of the transition relation.  The main benefit of this approach is
the simplicity of local termination measures in contrast to global ones. 
For instance, there are cases in which linear arithmetic suffices when using
local measures, while corresponding global measures require non-linear
functions or lexicographic orders.

%This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.

One drawback of the Ramsey-based approach is that the validity of the
termination argument relies on checking the \emph{transitive closure} of the
program, rather than a single step.  As such, there is experimental evidence
that most of the effort is spent in reachability
analysis~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/KroeningSTW10},
requiring the support of powerful safety checkers: there is a tradeoff
between the complexity of the termination arguments and that of checking
their validity.

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/HeizmannHP14}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.

While the summary  in Figure~\ref{fig:handletable} supports the observation
that the majority of existing termination analyses are designed for linear
programs and linear ranking functions, it also identifies another common
assumption made by state-of-the-art termination provers: that bit-vector
semantics and integer semantics give rise to the same termination behaviour. 
Thus, most of the techniques treat bit-vectors and floats as mathematical
integers and reals,
respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

@@

%% %\noindent {\bf Floats vs. reals.} 
%% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about
%% floating-point inherently difficult.
%% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
%% While the former does not terminate for reals, but does for floats, the latter 
%% terminates for reals, but does not for floats.

@@

We propose a general framework that uniformly computes lexicographic,
non-linear ranking functions supported by non-linear inductive invariants
for loops with arbitrary guards and transitions over bit-vectors and floats. 
It also computes non-linear recurrence sets, which comprise non-termination
proofs.  Our encoding uniformly multiple loops and nested loops, without
having to enumerate lassos.
%
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

@@

\item We designed a bit-level accurate technique for computing ranking
functions that correctly accounts for the wrap-around behavior caused by
under- and overflows in bit-vector and floating-point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

@@

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
To understand the issue, imagine the following loop with an non-deterministic assignment:

Only for certain values of \texttt{y} the loop is non-terminating.  In order
to restrict the values that may be assigned to \texttt{y}, we need an
assumption associated with the non-deterministic choice.  However, such an
assumption may cause a reachable state at the non-deterministic assignment
to not have any successor.  In such a case the termination formula would be
satisfiable, although the execution halts.  This situation can be avoided by
checking the existence of at least one successor for each reachable state at
the non-deterministic assignment.  Although we did put a reasonable amount
of effort into it, we failed to come up with a formulation for the
non-termination of such non-deterministic transition relations inside the
synthesis fragment.
%
%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.

@@

This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level. However, when dealing with nested loops, we need to compute the under-approximation of the inner loop, which requires a nested existential quantifier, 
resulting in the $\exists \forall \exists$ alternation. 
%% computing the under-approximation of the loop's body requires an extra quantifier alternation. 
%% The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
%% requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation. This situation is similar to the one encountered when proving non-termination of non-deterministic loops. 
To go around this limitation, we treat nested loops as a special case and give a separate formula for them in Figure~\ref{fig:nonterm-nested}. If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
\begin{itemize}
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
\end{itemize}

%% The issue appears when we try to under-approximate the body of the loop. 
%% Here we encounter a similar problem with the one for non-determinism mentioned above.

%% Consider the generic example in Figure~\ref{fig:nonterm-environment}, where we are attempting to prove non-termination
%% of the innermost loop $L_3$ by finding a closed recurrence set $N_3$. 
%% -- We will assume that the other two loops iterate only a finite number of times.

%% We now explain the constraints on the right of the figure.
%% Given that we prove non-termination, we try to  under-approximate the states from which $L_3$ can start executing. 
%% We do this in several steps:

%% $W_1$, $W_2$ and $W_3$ are inductive invariants corresponding to $L_1$, $L_2$ and $L_3$, respectively, such that:

%% - $W_1$ is an inductive invariant of $L_1$ that is strong enough to establish $W_2$ if $L_1$ exits (which it does by assumption).
%% - $W_2$ is an inductive invariant of $L_2$ that is strong enough to establish either $W_3$ or $N_3$ on entry to $L_3$, depending on whether $L_3$ terminates or not.
%% - Recall that $N_3$ is a recurrence set for $L_3$, so if a state $x \in N_3$ is reached, the loop $L_3$ will execute forever. Otherwise, $W_3$ holds on entry to $L_3$.
%% - $W_3$ is an inductive invariant of $L_3$ that is strong enough to re-establish $W_2$ when $L_3$ exits.

%% If the non-termination formula is satisfiable, $x_0$ is some initial state in which the program begins executing.

%% First, we under-approximate the states in which $L_2$ can start executing through $W_2$.
%% Then $N_1$ is an inductive invariant of $L_1$ that is strong
%% enough to establish $N_2$ if $L_1$ exits (which it does by assumption).  Now, $N_2$ is an inductive invariant
%% of $L_2$ that is strong enough to establish either $N_3$ or $N_4$ on entry to $L_3$.  Recall that $N_4$ is
%% a recurrence set for $L_3$, so if a state $x \in N_4$ is reached, the loop $L_3$ will execute forever.
%% Otherwise, $N_3$ holds on entry to $L_3$.  $N_3$ is an inductive invariant of $L_3$ that is strong enough to
%% re-establish $N_2$ when $L_3$ exits.

%% If the constraint system is satisfiable, $x_0$ is some initial
%% state in which the program begins executing.

%%   Since we assumed that $L_2$ iterated only a finite number of times,
%% $N_3$ must be established only a finite number of times, which means that $N_4$ will eventually be reached,
%% at which point $L_3$ will loop forever since $N_4$ is a recurrence set for $L_3$.

%% \begin{figure*}
%% \begin{framed}
%%  \begin{minipage}{0.19\textwidth}
%% \begin{lstlisting}
%% L1: while (c1) {
%%      B1;
%%     }

%% L2: while (c2) {
%%      P;

%% L3:  while (c3) {
%%        B3;
%%      }
%%     }
%% \end{lstlisting}
%% \end{minipage}
%% \vline
%% \begin{minipage}{0.82\textwidth}
%% \begin{align*}
%%  \exists W_1, W_2, W_3, N_3, x_0 . \forall x, x' . \\
%%   W_1(x_0) \, \wedge \\
%%   c_1(x) \wedge W_1(x) \wedge B_1(x, x') & \rightarrow W_1(x') \, \wedge \\
%%   \lnot c_1(x) \wedge W_1(x) & \rightarrow W_2(x) \, \wedge \\
%%   W_2(x) & \rightarrow c_2(x) \, \wedge \\
%%   W_2(x) \wedge P(x, x') & \rightarrow W_3(x') \vee N_3(x') \, \wedge \\
%%   c_3(x) \wedge W_3(x) \wedge B_3(x, x') & \rightarrow W_3(x') \, \wedge \\
%%   \lnot c_3(x) \wedge W_3(x) & \rightarrow W_2(x) \, \wedge \\
%%   N_3(x) & \rightarrow c_3(x) \, \wedge \\
%%   N_3(x) \wedge B_3(x, x') & \rightarrow N_3(x')
%% \end{align*}
%% \end{minipage}
%% \end{framed}

%% \caption{The constraint system for non-termination\label{fig:nonterm-environment}}
%% \end{figure*}

@@

Floating-point arithmetic:

>>>>

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
not large enough to prove termination for all terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

@@

\keywords
Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
Floating Point Ranking Functions.

@@

When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
This proof style proves that a transition relation is disjunctively well founded by finding a set of simple
functions, each of which proves well foundedness for part of the transition relation.
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve linear functions while corresponding global
measures require non-linear functions or lexicographic orders. 
%This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.

One difficulty with a Ramsey-based approach is that validity of the
termination argument can only be checked by considering the \emph{transitive
closure} of the program, rather than a single step.  As such, there is
experimental evidence that most of the time is spent in reachability
analysis \cite{DBLP:conf/pldi/CookPR06}, requiring the support of powerful
safety checkers: there is a tradeoff between the complexity of the
termination arguments and that of checking their validity.

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its extreme: \emph{unrestricted} termination arguments.  This
means that our ranking functions may involve non-linearity and lexicographic
orders (we do not commit to any particular syntactic form, and do not use
templates).

While the summary  in Figure~\ref{fig:handletable} supports the observation
that the majority of existing termination analyses are designed for linear
programs and linear ranking functions, it also identifies another common
assumption made by state of the art termination provers: that bit-vector
semantics and integer semantics give rise to the same termination behaviour. 
Thus, most of the techniques treat bit-vectors and floats as mathematical
integers and reals,
respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

@@

%% %\noindent {\bf Floats vs. reals.} 
%% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
%% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
%% While the former does not terminate for reals, but does for floats, the latter 
%% terminates for reals, but does not for floats.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
lexicographic, non-linear 
ranking functions supported by non-linear inductive invariants for loops with 
arbitrary guards and transitions over bit-vectors and floats.
It also computes non-linear recurrence sets, which comprise non-termination proofs.
Our encoding uniformly multiple loops and nested loops, without having to enumerate
lassos.
%
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

@@

\item We designed a bit-level accurate technique for computing ranking
functions that correctly accounts for the wrap-around behavior caused by
under- and overflows in bit-vector and floating point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

@@

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

@@

This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level.

However, when dealing with nested loops, we need to  under-approximate the inner loop, requiring a nested existential quantifier, 
resulting in the $\exists \forall \exists$ alternation.  At first sight, this seems fatal for our formulation.  Luckily, it is possible to
reformulate non-termination as an $\exists \forall$ formula by observing that, unlike a ranking function, the defining property of a
recurrence set is \emph{non relational} -- if we end up in the recurrence set, we don't care exactly where we came from as long as
we know that it was also somewhere in the recurrence set.  This allows us to cast non-termination of nested loops as the formula
shown in Figure~\ref{fig:nonterm-nested}.

If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
\begin{itemize}
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
\item If $L_2$ is non-terminating, then $N_2 \wedge G_2$ is its recurrence set.
\end{itemize}

@@

Floating point arithmetic:
